Relational programming in Smalltalk

### 2019-03-13 booklet/article.tex
The selector Object>>#cons: now double dispatches to gain more flexibility.

\newcommand{\stcode}[1]{\begin{minted}[fontsize=\footnotesize]{smalltalk} #1 \end{minted}}

@@

\begin{abstract}
This booklet is an extended description of $\mu$Kanren, a relational
interpreter formerly implemented in Scheme and ported in Smalltalk for the sake
of understanding and, of course, fun. It enjoys simplicity and elegance despite
the complex nature of logic systems; moreover, using an uniform but powerful
language such as Smalltalk it is possible to appreciate and benefit from both
of them.
\end{abstract}

@@

This document extends the presentation of $\mu$Kanren shown at the last ESUG
conference \citep{ESUG2018} in two ways: (i)~it recaps existing ``literature''
on the subject and (ii)~it describes in more details our implementation coded
in Smalltalk.

It is a waste of time to repeat ourself on $\mu$Kanren's definition; on the
contrary, we believe interesting to look up and to watch over the many
implementations it has inspired since its seminal paper \citep{Hemann:muKanren}
and how programmers stretch the canonical implementation to take advantage of
their favorite programming language's features.

Certainly the book \citep{Friedman:Reasoned:Schemer} is the main reference
written in the same style of the other \emph{little books} of which Dan
Friedman coauthors them all; moreover, Byrd's dissertation \citep{Byrd:PhD}
shows advanced stuffs and techniques while the repository
\url{http://minikanren.org/} -- updated regularly by Byrd himself -- keeps
pointers to implementations, articles and presentations. All these references
use the Scheme programming language, $\mu$Kanren is \emph{purely functional}
therefore any language that has the same power of the $\lambda$-calculus is a
candidate for implementing it, Scheme in particular; in parallel, the
implementation with symbolic constraints has the same nature but uses Scheme
macrology and in some aspect it requires little effort to port to other
languages that haven't macros.

Another implementation using the \emph{Clojure} dialect of Lisp is
\citep{clojure:core.logic} which targets the JVM and provides constraints and
nominal logic programming; on the other hand, \citep{sullivan:microkanrenhs}
is a clean and concise implementation using Haskell, while many others simpler
versions can be found, written for the sake of understanding and, ultimately,
for fun.

Finally, your humble authors believe that $\mu$Kanren is a sophisticated
\textit{educational machine} because it concerns and mixes together unification,
lazy streams, backtracking and optimization techniques; moreover, it allows
programmers to write their own version according to their personal taste and
style, ranging from implementations pretty imperative to ones built over
categories.  We repeat the exercise of writing it in \citep{Nocentini:kanrens}
targeting different languages, in particular
\begin{itemize}
\item using \emph{Python} we take advantage of native \verb|generator|s objects
and we provide a \emph{fair} enumeration strategy that the canonical one
hasn't; on the contrary, we experience computational limits because of
\texttt{RecursionError} exceptions. This implementation is complete and
parallels the one shown in the reference book, accompanied by a unit test suite
that covers \emph{all} questions explained in the book and some new
applications such as (i)~finding fixed points of a set of inference rules and
enumeration of classes of combinatorial objects.
\item using \emph{OCaml} the syntax of the language can be cleanly defined by
algebraic datatypes and the type system helps in the design of the required
operators.  While object-oriented languages make extensibility a relative
simple task, this implementation is harder unless we can make changes to basic
types, at least for what concerns the central \emph{unification} procedure.
All of this is preparatory to modify the tactics mechanism of the \emph{HOL
Light} theorem prover in order to introduce the relational paradigm in that
environment.
\end{itemize}
The current work describes a Smalltalk implementation of the basic system
with the aim to learn and experiment with a pure object environment, such as 
Pharo and Squeak. Starting small, the next example we show three different
ways to express equality between \verb|SmallInteger| objects:

\begin{example}
In math a relation $P$ is usually characterized by a statement such as
\begin{displaymath}
\forall a,b,c.\,P(a,b,c) \leftrightarrow a + b = c \quad\text{from which}\quad
P(1,2,3) \quad\text{can be proved.}
\end{displaymath}
The same check can be expressed in Smalltalk using either the
\textit{imperative style}
\begin{minted}[fontsize=\footnotesize]{smalltalk}
a := 1.  b := 2.  c := a + b.  Object assert: [ c = 3 ]
\end{minted}
or the \textit{functional style}
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Object assert: [ ([ :a :b | a + b ] value: 1 value: 2) = 3 ]
\end{minted}
or, finally, the \textit{declarative style}
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Object assert: [ [ :a :b :c | a + b = c ] value: 1 value: 2 value: 3 ]
\end{minted}
which closely resembles the mathematical statement.
\end{example}

@@

@@

\begin{example}
Let $x$ and $y$ be free variables, the set
$$\lbrace cons(x,cons(x,nil)) = cons(2,y)\rbrace$$
has solution $\theta = \lbrace x \mapsto 2, y \mapsto cons(2,nil) \rbrace$;
on the contrary, the set
$$ \lbrace y = cons(2,y) \rbrace $$
has no \textit{finite} solution, nonetheless
$$\theta = \lbrace y \mapsto cons(2,cons(2,cons(2,...))) \rbrace$$
is a solution upto \textit{bisimulation}
\citep{10.1007/BFb0017309, DBLP:books/daglib/0067019}.
\end{example}
\section{Conclusions}

A quine generator is described in \citep{Byrd:2012:MLU:2661103.2661105} and
we aim to reproduce it.

>>>>

\newcommand{\stcode}[1]{\mintinline[fontsize=\footnotesize]{smalltalk}{#1}}

@@

\begin{abstract}
This booklet is an extended description of $\mu$Kanren, a relational
interpreter formerly implemented in Scheme and ported in Smalltalk for the sake
of understanding and, of course, fun. It enjoys simplicity and elegance despite
the complex nature of logic systems; moreover, using an uniform but powerful
language such as Smalltalk it is possible to appreciate and benefit from both
properties.
\end{abstract}

@@

\input{introduction.tex}

@@

\section{Preparing the wires...}

\subsection{Goals and substitutions}

A \textit{goal} is any object that responds to the selector \Verb|#onState:| so
that it receives a substitution and produces an ordered sequence of zero or
more substitutions. In turn, a substitution is an unordered collection of
values which are indexed by \textit{logic variables}, therefore a
\Verb|Dictionary| object fits perfectly; we wrap it in a \Verb|State| object
that also keeps record of an incremental numbering \Verb|birthdate| whose
meaning will be clarified later when we'll describe the \Verb|Fresh|
goal\footnote{put here the reference to its subsection}. An empty substitution
is created by \stcode{Dictionary new asState} where
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Dictionary>>#asState
    ^ State new 
        birthdate: 0;
        substitution: self;
        yourself
\end{minted}
and it can be augmented by
\begin{minted}[fontsize=\footnotesize]{smalltalk}
State>>#at: aVar put: aValue 
    | s |
    s := substitution copy.
    s "check that if a key already exists at aVar, then the values must be the same"
      at: aVar
      ifPresent: [:v | aValue = v ifFalse: [UnificationError signal]]
      ifAbsentPut: aValue.
    ^ self class new 
        birthdate: birthdate;
        substitution: s;
        yourself
\end{minted}
mantaining a protocol coherent to the one of \Verb|Dictionary| objects for
extension.

\subsection{Chains}

Because the sequence of substitutions may be infinite, we
represent it not as a \Verb|LinkedList| object but as a \Verb|Chain| object;
as usual, we find \Verb|BlockClosure| objects good candidates to represent (a possibly
infinite) computation. Some examples are in order,
\begin{minted}[fontsize=\footnotesize]{smalltalk}
ChainTest>>#fibs
    ^ self fib: 0 fib: 1

ChainTest>>#fib: m fib: n 
    ^ [self fib: n fib: m + n] links: m

ChainTest>>#gcd: m gcd: n 
    ^ [n isZero 
            ifTrue: [Chain with: m]
            ifFalse: [self gcd: n gcd: m \\ n]] 
        links: m

ChainTest>>#nats
    ^ self ints: 0

ChainTest>>#ints: i 
    ^ [self ints: i + 1] links: i

ChainTest>>#upto2
    ^ [[[Chain bottom] links: 2] links: 1] links: 0
\end{minted}
where blocks define the message \Verb|#links:| as 
\begin{minted}[fontsize=\footnotesize]{smalltalk}
BlockClosure>>#links: anObj
    ^ Chain item: anObj linker: self
\end{minted}
exposing the abstract class \Verb|Chain| and its class-side messages
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Chain class>>#bottom
    ^ Bottom new

Chain class>>#item: anObj linker: aBlockClosure 
    ^ Knot new 
        item: anObj;
        linker: aBlockClosure;
        yourself

Chain class>>#with: anObj 
    ^ [Chain bottom] links: anObj

Chain class>>#repeat: anObj 
    ^ [self repeat: anObj] links: anObj
\end{minted}
together with its two subclasses \Verb|Knot| and \Verb|Bottom|; a comprehensive
test case follows
\begin{minted}[fontsize=\footnotesize]{smalltalk}
ChainTest>>#testNumbers
    self 
        assert: Chain bottom contents 
        equals: LinkedList new.
    self
        assert: (self nats next: 10) contents
        equals: (0 to: 9).
    self
        assert: (self upto2 next: 10) contents
        equals: (0 to: 2).
    self
        assert: (self upto2 next: 1) contents
        equals: (0 to: 0).
    self 
        assert: (self fibs next: 10) contents 
        equals: {0. 1. 1. 2. 3. 5. 8. 13. 21. 34}.
    self 
        assert: ((self collatz: 10) next: 10) contents 
        equals: {10. 5. 16. 8. 4. 2. 1. 4. 2. 1}.
    self 
        assert: ((self gcd: 18 gcd: 32) next: 10) contents 
        equals: {18. 32. 18. 14. 4. 2. 2}.
    self 
        assert: ((Chain repeat: 4) next: 10) contents 
        equals: {4. 4. 4. 4. 4. 4. 4. 4. 4. 4}
\end{minted}
On one hand, when a \Verb|Chain| object (i)~receives the selector
\Verb|#next:| that carries a \Verb|SmallInteger| object $i$ it keeps at most
$i$ objects and then truncate itself, precisely  
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Bottom>>#next: anInteger 
    ^ self

Knot>>#next: n 
    ^ n isZero
        ifTrue: [Chain bottom]
        ifFalse: [[self next value next: n - 1] links: item]
\end{minted}
On the other hand, when a \Verb|Chain| object (ii)~receives the selector
\Verb|#contents| it builds a \Verb|LinkedList| object \textit{eagerly},
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Bottom>>#contents
    ^ LinkedList new

Knot>>#contents
    ^ self next contents 
        addFirst: item;
        yourself
\end{minted}
where \Verb|Knot| objects provide the following simple but fundamental message
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Knot>>#next
    ^ linker value
\end{minted}
Observe that the \Verb|Chain| hierarchy support part of the protocol provided
by\Verb|ReadStream| objects because of the messages \Verb|#next|, \Verb|#next:|
and \Verb|#contents|.

@@

\begin{example}
Let $x$ and $y$ be free variables, the set
$$\lbrace cons(x,cons(x,nil)) = cons(2,y)\rbrace$$
has solution $\theta = \lbrace x \mapsto 2, y \mapsto cons(2,nil) \rbrace$;
on the contrary, the set
$$ \lbrace y = cons(2,y) \rbrace $$
has no \textit{finite} solution, nonetheless
$$\theta = \lbrace y \mapsto cons(2,cons(2,cons(2,...))) \rbrace$$
is a solution upto \textit{bisimulation}
\citep{10.1007/BFb0017309, DBLP:books/daglib/0067019}.
\end{example}
\section{Conclusions and further works}

\begin{itemize}
\item A quine generator is described in \citep{Byrd:2012:MLU:2661103.2661105}
and we aim to reproduce it.

\item using \emph{OCaml} the syntax of the language can be cleanly defined by
algebraic datatypes and the type system helps in the design of the required
operators.  While object-oriented languages make extensibility a relative
simple task, this implementation is harder unless we can make changes to basic
types, at least for what concerns the central \emph{unification} procedure.
All of this is preparatory to modify the tactics mechanism of the \emph{HOL
Light} theorem prover in order to introduce the relational paradigm in that
environment.
\end{itemize}
